Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-01-09 ms.tex
editing intro

In order to understand why most SMBHs appear to be inactive, it is
necessary to understand their gaseous environments.  Hot gas near the
SMBH sphere of influence (the `circumnuclear medium' or CNM) controls
the central mass accretion rate $\dot{M}$.  Better knowledge of how
$\dot{M}$ varies with the SMBH mass $\Mbh$ and other properties of the
galactic nucleus is necessary to understand the co-evolution of SMBHs
and their host galaxies with cosmic time.  SMBH growth in the low
redshift universe, for instance, appears to be dominated by low mass
black holes (e.g.~\citealt{Heckman+04}).  Although this is often
attributed to cosmological ``down sizing" (e.g, \citealt{Gallo+08}),
the physical processes involved may be distinct from those operating
at higher SMBH masses and accretion rates in quasars.

The accretion rates of low mass SMBHs and their resulting nuclear
X-ray luminosities $L_X$ also have implications for observational
constraints on their occupation fraction in low mass galaxies and the
possible existence of intermediate mass black holes.
\citet{MillerGallo+:2014a} combine the relationship between $\Mbh$ and
$L_X$ from a sample of relatively high mass galaxies
(\citealt{Gallo+08}) with upper limits on $L_X$ in several lower mass
galaxies, to infer tentative evidence for an occupation fraction less
than unity.  However, this method relies on extrapolating a power-law
fit to the $L_X-\Mbh$ distribution to low $\Mbh$, below the
instrumental detection threshold.  The conclusions of
\citet{MillerGallo+:2014a} are thus sensitive to possible changes in
the $L_X$-$\Mbh$ relationship at small $\Mbh$.
% AG: In our model we obtain a steeper relationship, but we have difficulty matching the observed relationship above the detection threshold

The density profile of the CNM also influences the emission from
stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11};
\citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This event was powered by a transient relativistic jet, which produced
synchrotron radio emission as the jet material was decelerated by
shock interaction with the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of {\it Swift} J1644+57 showed that the CNM density was
significantly lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in {\it Swift}
J1644+57.  Variations in the properties of the CNM could thus in
principle help explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

Gas that forms the CNM of a SMBH may originate from several sources
(\citealt{Ho:2009a}): (1) wind mass loss from (predominantly evolved)
stars; (2) star-star collisions; (3) the unbound debris from TDEs.
Stellar wind mass loss is probably the dominant source in most cases
insofar as collisions are relevant only in very dense stellar
environments for young stellar populations (\citealt{Rubin&Loeb11}),
while TDEs are estimated to provide a subdominant contribution to the
time-averaged accretion rate of quiescent SMBHs
(\citealt{MacLeod+13}).

\citet{Ho:2009a} calculates the accretion rates onto a sample of
early-type galaxies both by using X-ray observations to determine the
Bondi accretion rate, and by estimating the contributions from the
mass loss of evolved stars.  Both methods lead him to conclude that
the available reservoir of gas is more than sufficient to power
low-luminosity AGN, assuming the standard radiative efficiency $\eta
\sim 0.1$ for thin disk accretion.  Indeed, there is now strong
observational evidence that the low levels of AGN activity in
quiescent galaxies results from accretion proceeding in a radiatively
inefficient mode (\citealt{Yuan&Narayan14}), either due to the
gravitationally-released energy being advected across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to only a fraction of the gas
inflowing on large scales ultimately reaching the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

The accretion rates of SMBHs can in principle be estimated directly by
assuming Bondi accretion and using {\it Chandra} X-ray observations to
estimate the gas temperature and density near the Bondi radius.  This
technique has led to the intriguing result that the Bondi accretion
power correlates with the AGN jet power (e.g.,
\citealt{AllenDunn+:2006a}; \citealt{Russell+13};
\citealt{FujitaKawakatu+:2014a}), the latter estimated from the energy
required to inflate the observed cavities within the X-ray emitting
gas.  It is, however, almost never possible to resolve the Bondi
radius, thus requiring all such studies to extrapolate the observed
gas temperature and density profiles to much smaller radial scales.
%In fact, the temperature profile is assumed to be flat inside of the
%Bondi radius.  In reality, the cusp in the velocity dispersion near
%the SMBH, should cause a cusp in the gas temperature profile (assuming
%the kinetic energy of stellar winds is efficiently thermalized in
%shocks).
%% AG-give sense of scale.

Another approach to determine the SMBH accretion rate, which we adopt
in this paper, is to directly calculate the density, velocity and
temperature profiles of the CNM using a physically motivated
hydrodynamic model.  Mass is injected to the nuclear environment via
stellar winds, while energy input results from a number of sources,
such as wind shock heating and supernovae (e.g.,
\citealt{Quataert:2004a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}).
Unlike previous works, however, which focus on modeling the nuclei of
individual galaxies, here we systematically analyze the properties of
the CNM across a large sample of quiescent galaxies with a range of
SMBH masses and other properties (\citealt{WangMerritt:2004a};
hereafter WM04).  For each galaxy, stellar winds deposit mass and
energy across a range of radii calculated directly from the observed
{\it stellar} density profile.

The one-dimensional steady state flow of gas is usually characterized
by an inflow-outflow structure, with a critical radius known as the
``stagnation radius" $\rs$ where the radial velocity goes to zero.
Mass loss from stars interior to the stagnation radius is accreted,
while that outside $\rs$ is unbound in an outflow from the nucleus.
The stagnation radius, rather than the Bondi radius, is thus
ultimately responsible for setting the SMBH accretion rate, although
$\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion. 
The main goal of our analysis is to provide a framework for
interpreting the properties of quiescent galactic nuclei given the
comparatively well understood physics of stellar winds and supernova
heating.

@@

>>>>

To understand why most SMBHs appear to be inactive requires understanding their gaseous environments.  Hot gas near the
SMBH sphere of influence (hereafter the `circumnuclear medium', or CNM) controls the central mass accretion rate $\dot{M}$.  The latter in turn affects the luminosity of the accretion flow and the possible impact of its energy and momentum input (`feedback') on larger scales.  The presence of dense gas in the nucleus can also lead to runaway cooling instabilities, resulting in bursty star-formation or episodic AGN activity (reference Ciotti and others).

Knowledge of how $\dot{M}$ varies with the SMBH mass $\Mbh$ and other properties of the
galactic nucleus informs key questions related to co-evolution of SMBHs and their host galaxies with cosmic time.  SMBH growth in the low redshift universe appears to be dominated by lower mass black holes (e.g.~\citealt{Heckman+04}).  This is often
attributed to cosmological `down sizing' (e.g, \citealt{Gallo+08}), even though the physical processes involved may be distinct from those operating at higher SMBH masses and accretion rates in quasars.

The accretion rates of low mass SMBHs, and their resulting nuclear X-ray luminosities $L_X$, also have implications for observational
constraints on their occupation fraction in low mass galaxies and the
possible existence of intermediate mass black holes.  By combining the average relationship between $\Mbh$ and
$L_X$ measured from a sample of high mass galaxies (\citealt{Gallo+08}) with upper limits on $L_X$ in lower mass
galaxies, \citet{MillerGallo+:2014a}  infer tentative evidence for an occupation fraction less than unity in the latter.  This method, however, relies on extrapolating a power-law fit to the $L_X-\Mbh$ distribution to low $\Mbh$ below the detection threshold.

%The conclusions of
%\citet{MillerGallo+:2014a} are thus sensitive to possible changes in
%the $L_X$-$\Mbh$ relationship at small $\Mbh$.
% AG: In our model we obtain a steeper relationship, but we have difficulty matching the observed relationship above the detection threshold

The gas density of the CNM also influences the emission from stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11}; \citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This event was powered by a transient relativistic jet, which produced
synchrotron radio emission as the jet material was decelerated by
shock interaction with the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of J1644+57 showed that the CNM density was
significantly lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in J1644+57.  Variations in the properties of the CNM could thus in
principle help explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

Gas comprising the CNM can in principle originate from a variety of sources: (1) wind mass loss from (predominantly evolved) stars; (2) star-star collisions; and (3) unbound debris from TDEs.  Stellar wind mass loss is probably the dominant source insofar as collisions are relevant only in very dense stellar environments for young stellar populations (\citealt{Rubin&Loeb11}), while \citet{MacLeod+13} find that TDEs contribute subdominantly to the time-averaged accretion rate of quiescent SMBHs.

\citet{Ho:2009a} estimates the accretion rates onto a sample of early-type galaxies both by using X-ray observations to determine the Bondi accretion rate and using estimates for the mass loss rate of evolved stars.  Both methods lead to the conclusion that the available gas reservoir is more than sufficient to power the observed low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative efficiency for thin disk accretion.  Several independent strains of evidence now suggest that low-luminosity AGN result from accretion proceeding in a radiatively inefficient mode (\citealt{Yuan&Narayan14}), due either to the advection of gravitationally-released energy across the SMBH horizon (e.g.~\citealt{Narayan&Yi95}) or due to a low efficiency for the gas inflowing on large scales ultimately reaching the SMBH (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

%In fact, the temperature profile is assumed to be flat inside of the
%Bondi radius.  In reality, the cusp in the velocity dispersion near
%the SMBH, should cause a cusp in the gas temperature profile (assuming
%the kinetic energy of stellar winds is efficiently thermalized in
%shocks).
%% AG-give sense of scale.

Another approach to determine the SMBH accretion rate, which we adopt in this paper, is to directly calculate the density, velocity and temperature profiles of the CNM using a physically motivated
hydrodynamic model.  Mass is injected to the nuclear environment via
stellar winds, while energy input results from several sources including stellar winds, supernovae, and AGN feedback (e.g., \citealt{Quataert:2004a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}).  Unlike previous works focusing on modeling individual galaxies, here we systematically analyze the properties of the CNM across a sample of galaxies with a range of SMBH masses, stellar density profiles, and star formation histories (\citealt{WangMerritt:2004a};
hereafter WM04).

Many past studies have focused on relatively massive galaxies.  A key result of this work is the onset of , and the need for AGN feedback.  Here we instead neglect radiative cooling, focusing on time-dependent models.  This simple approach allows us to systematically and to derive simple analytic expressions that prove useful in exploring their sensitivity on.

The one-dimensional steady state flow of gas is usually characterized
by an inflow-outflow structure, with a critical radius known as the
``stagnation radius" $\rs$ where the radial velocity goes to zero.
Mass loss from stars interior to the stagnation radius is accreted,
while that outside $\rs$ is unbound in an outflow from the nucleus.
The stagnation radius, rather than the Bondi radius, is thus
ultimately responsible for setting the SMBH accretion rate, although (when it exists)
$\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion. 
The main goal of our analysis is to provide a framework for
interpreting the properties of quiescent galactic nuclei given the
comparatively well understood physics of stellar winds and supernova
heating.

@@

\subsection{TDE Jets}

Plot showing final lorentz factor after the reverse shock crossing as a function of BH mass, using average star formation history.  place J1644+57 on the plot.

\subsection{Jet Power Correlation}

Accretion rates onto SMBHs can in principle be estimated directly by assuming Bondi accretion and using high resolution X-ray observations to estimate the gas temperature and density near the Bondi radius.  This technique has led to the intriguing conclusion that the Bondi accretion power correlates with the AGN jet power (e.g.,
\citealt{AllenDunn+:2006a}; \citealt{Russell+13};
\citealt{FujitaKawakatu+:2014a}), the latter estimated from the energy
required to inflate the observed cavities within the X-ray emitting
gas.  In practice is it almost never possible to resolve the Bondi radius, therefore requiring all such studies to extrapolate the observed gas temperature and density profiles to smaller radial scales.
